---
title: "R-GIS_tutorial.Rmd"
author: "Robert A. Stevens"
date: "September 14, 2015"
output: html_document
---

From https://github.com/Pakillo/R-GIS-tutorial

# Spatial data in R: Using R as a GIS     

A tutorial to perform basic operations with spatial data in R, such as importing and exporting data (both vectorial and raster), plotting, analysing and making maps.

Francisco Rodriguez-Sanchez http://sites.google.com/site/rodriguezsanchezf

v 2.2  

27-01-2015 

Licence: CC BY 4.0 http://creativecommons.org/licenses/by/4.0/

Check out code and latest version at GitHub https://github.com/Pakillo/R-GIS-tutorial/blob/master/R-GIS_tutorial.md

## CONTENTS

1.0 INTRODUCTION  
2.0 GENERIC MAPPING  
2.1 Retrieving base maps from Google with `gmap` function in package `dismo`  
2.2 `RgoogleMaps`: Map your data onto Google Map tiles  
2.3 `googleVis`: visualise data in a web browser using Google Visualisation API  
2.4 `RWorldMap`: mapping global data  
3.0 SPATIAL VECTOR DATA (points, lines, polygons)  
3.1 Example dataset: retrieve point occurrence data from GBIF  
3.2 Making data 'spatial'  
3.3 Define spatial projection  
3.4 Quickly plotting point data on a map  
3.5 Subsetting and mapping again  
3.6 Mapping vectorial data (points, polygons, polylines)  
3.7 Drawing polygons and polylines (e.g. for digitising)  
3.8 Converting between formats, reading in, and saving spatial vector data  
3.9 Changing projection of spatial vector data  
4.0 USING RASTER (GRID) DATA  
4.1 Downloading raster climate data from internet  
4.2 Loading a raster layer  
4.3 Creating a raster stack  
4.4 Raster bricks  
4.5 Crop rasters  
4.6 Define spatial projection of the rasters  
4.7 Changing projection  
4.8 Plotting raster data  
4.9 Spatial autocorrelation  
4.10 Extract values from raster  
4.11 Rasterize vector data (points, lines or polygons)  
4.12 Changing raster resolution  
4.13 Spline interpolation  
4.14 Setting all rasters to the same extent, projection and resolution all in one  
4.15 Elevations, slope, aspect, etc  
4.16 Saving and exporting raster data  
5.0 SPATIAL STATISTICS  
5.1 Point pattern analysis  
5.2 Geostatistics  
6.0 INTERACTING WITH OTHER GIS  
7.0 OTHER USEFUL PACKAGES  
8.0 TO LEARN MORE

# 1. INTRODUCTION

R is great not only for doing statistics, but also for many other tasks, including GIS analysis and working with spatial data. For instance, R is capable of doing wonderful maps such as [this](http://spatialanalysis.co.uk/wp-content/uploads/2012/02/bike_ggplot.png) or [this](http://oscarperpinan.github.io/spacetime-vis/images/airMadrid_stamen.png). In this tutorial I will show some basic GIS functionality in R.

### Basic packages

```{r message=FALSE}
# and their dependencies
library(sp)         # classes for spatial data
library(raster)     # grids, rasters
library(rasterVis)  # raster visualisation
library(maptools)   
library(rgeos)
library(rgdal)
library(XML)
library(dismo)
library(RgoogleMaps) 
library(googleVis) 
library(rworldmap) # For a short introduction type: vignette('rworldmap')
library(dismo)      # check also the nice "rgbif" package! 
library(jsonlite)
#library(rworldmap) # library rworldmap provides different types of global maps, e.g:
#require(RgoogleMaps)
library(gtools)
library(rgl)
library(fields)
```

There are many other useful packages, e.g. check [CRAN Spatial Task View](http://cran.r-project.org/web/views/Spatial.html). Some of them will be used below.
     
# 2. GENERIC MAPPING

## 2.1 Retrieving base maps from Google with `gmap` function in package `dismo`

Some examples:

Getting maps for countries:

```{r gmap1, message=FALSE}
mymap <- gmap("France")   # choose whatever country
plot(mymap)
```

Choose map type:

```{r gmap2, message=FALSE}
mymap <- gmap("France", type = "satellite")  
plot(mymap)
```

Choose zoom level:

```{r gmap3, message=FALSE}
mymap <- gmap("France", type = "satellite", exp = 3)
plot(mymap)
```

Save the map as a file in your working directory for future use

```{r message=FALSE}
mymap <- gmap("France", type = "satellite", filename = "France.gmap")    
```

Now get a map for a region drawn at hand

```{r eval=FALSE}
mymap <- gmap("Europe")
plot(mymap)
select.area <- drawExtent()   
# now click 2 times on the map to select your region
mymap <- gmap(select.area)
plot(mymap)
# See ?gmap for many other possibilities
```

## 2.2 `RgoogleMaps`: Map your data onto Google Map tiles

Get base maps from Google (a file will be saved in your working directory)

```{r message=FALSE, results='hide'}
newmap <- GetMap(center = c(36.7, -5.9), zoom = 10, destfile = "newmap.png", 
                 maptype = "satellite")   
# Now using bounding box instead of center coordinates:
newmap2 <- GetMap.bbox(lonR = c(-5, -6), latR = c(36, 37), destfile = "newmap2.png", 
                       maptype="terrain")   
# Try different maptypes
newmap3 <- GetMap.bbox(lonR = c(-5, -6), latR = c(36, 37), destfile = "newmap3.png", 
                       maptype="satellite")
```

Now plot data onto these maps, e.g. these 3 points

```{r}
PlotOnStaticMap(lat = c(36.3, 35.8, 36.4), lon = c(-5.5, -5.6, -5.8), zoom = 10, 
                cex = 4, pch = 19, col = "red", FUN = points, add = F, size = c(640, 640))
```

## 2.3 `googleVis`: visualise data in a web browser using Google Visualisation API

Run `demo(googleVis)` to see all the possibilities

```{r setOptions, echo=FALSE}
op <- options(gvis.plot.tag = "chart")  
# necessary so that googleVis works with knitr, 
# see http://lamages.blogspot.co.uk/2012/10/googlevis-032-is-released-better.html
```

### Example: plot country-level data

```{r results='asis', tidy=FALSE, eval=TRUE}
data(Exports)    # a simple data frame
Geo <- gvisGeoMap(Exports, locationvar = "Country", numvar = "Profit", 
                  options = list(height=400, dataMode = 'regions'))
plot(Geo)
```

Using `print(Geo)` we can get the HTML code to embed the map in a web page!

### Example: Plotting point data onto a google map (internet)

```{r results='asis', tidy=FALSE, eval=TRUE}
data(Andrew)
M1 <- gvisMap(Andrew, "LatLong", "Tip", 
              options = list(showTip = TRUE, showLine = F, enableScrollWheel = TRUE, 
                             mapType = 'satellite', useMapTypeControl = TRUE, width = 800,
                             height = 400))
plot(M1)
```

## 2.4 `RWorldMap`: mapping global data

Some examples

```{r message=FALSE, warning=FALSE}
newmap <- getMap(resolution = "coarse")    # different resolutions available
plot(newmap)
```

```{r message=FALSE} 
mapCountryData()
```

```{r message=FALSE}
mapCountryData(mapRegion = "europe")
```

```{r message=FALSE}
mapGriddedData()
```

```{r message=FALSE}
mapGriddedData(mapRegion = "europe")  
```

# 3. SPATIAL VECTOR DATA (points, lines, polygons)

## 3.1 Example dataset: retrieve point occurrence data from GBIF

Let's create an example dataset: retrieve occurrence data for the laurel tree (Laurus nobilis) from the [Global Biodiversity Information Facility (GBIF)](http://gbif.org)

```{r message=FALSE}
laurus <- gbif("Laurus", "nobilis")      
# get data frame with spatial coordinates (points)
locs <- subset(laurus, select = c("country", "lat", "lon"))
head(locs)    # a simple data frame with coordinates
# Discard data with errors in coordinates:
locs <- subset(locs, locs$lat < 90)
```

## 3.2 Making data 'spatial'

So we have got a simple dataframe containing spatial coordinates. Let's make these data explicitly *spatial*

```{r}
coordinates(locs) <- c("lon", "lat")    # set spatial coordinates
plot(locs)
```

## 3.3 Define spatial projection

Important: define geographical projection. 

Consult the appropriate PROJ.4 description here: 

[http://www.spatialreference.org/](http://www.spatialreference.org/)

```{r}
crs.geo <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")    # geographical, datum WGS84
proj4string(locs) <- crs.geo     # define projection system of our data
summary(locs)
```

## 3.4 Quickly plotting point data on a map

```{r}
plot(locs, pch = 20, col = "steelblue")
data(coastsCoarse) 
data(countriesLow)
plot(coastsCoarse, add = TRUE)
```

## 3.5 Subsetting and mapping again

```{r}
table(locs$country)     # see localities of Laurus nobilis by country
locs.gb <- subset(locs, locs$country == "United Kingdom")    # select only locs in UK
plot(locs.gb, pch = 20, cex = 2, col = "steelblue")
title("Laurus nobilis occurrences in UK")
plot(countriesLow, add = TRUE)
summary(locs.gb)
```

## 3.6 Mapping vectorial data (points, polygons, polylines)

### Mapping vectorial data using `gmap` from `dismo`

```{r}
gbmap <- gmap(locs.gb, type = "satellite")
locs.gb.merc <- Mercator(locs.gb)    # Google Maps are in Mercator projection. 
# This function projects the points to that projection to enable mapping
plot(gbmap)
points(locs.gb.merc, pch = 20, col = "red")
```

### Mapping vectorial data with `RgoogleMaps` 

```{r message=FALSE}
locs.gb.coords <- as.data.frame(coordinates(locs.gb))    # retrieves coordinates 
# (1st column for longitude, 2nd column for latitude)
PlotOnStaticMap(lat = locs.gb.coords$lat, lon = locs.gb.coords$lon, 
                zoom = 5, cex = 1.4, pch = 19, col = "red", FUN = points, add = FALSE,
                size = c(640, 640))
```

Download base map from Google Maps and plot onto it

```{r message=FALSE}
map.lim <- qbbox (locs.gb.coords$lat, locs.gb.coords$lon, TYPE = "all")    
# define region of interest (bounding box)
mymap <- GetMap.bbox(map.lim$lonR, map.lim$latR, destfile = "gmap.png", 
                     maptype ="satellite")
# see the file in the wd
PlotOnStaticMap(mymap, lat = locs.gb.coords$lat, lon = locs.gb.coords$lon, 
                zoom = NULL, cex = 1.3, pch = 19, col = "red", FUN = points, add = FALSE,
                size = c(640, 640))
```

Using different background (base map)

```{r message=FALSE}
mymap <- GetMap.bbox(map.lim$lonR, map.lim$latR, destfile = "gmap.png", maptype = "hybrid")
PlotOnStaticMap(mymap, lat = locs.gb.coords$lat, lon = locs.gb.coords$lon, 
                zoom = NULL, cex =1.3, pch = 19, col ="red", FUN = points, add = FALSE,
                size = c(640, 640))
```

### Map vectorial data with `googleVis` (internet) 

```{r results='asis', tidy=FALSE, eval=TRUE}
points.gb <- as.data.frame(locs.gb)
points.gb$latlon <- paste(points.gb$lat, points.gb$lon, sep = ":")
map.gb <- gvisMap(points.gb, locationvar = "latlon", tipvar = "country", 
                  options = list(showTip = TRUE, showLine = FALSE, 
                                 enableScrollWheel = TRUE, useMapTypeControl = TRUE, 
                                 width = 1400, height = 800))
plot(map.gb)
#print(map.gb)    # get HTML suitable for a webpage
```

## 3.7 Drawing polygons and polylines (e.g. for digitising)

```{r eval=FALSE}
plot(gbmap)
mypolygon <- drawPoly()    # click on the map to draw a polygon and press ESC when finished
summary(mypolygon)    # now you have a spatial polygon! Easy, isn't it?
```

## 3.8 Converting between formats, reading in, and saving spatial vector data

### Exporting KML (Google Earth)

```{r}
writeOGR(locs.gb, dsn = "locsgb.kml", layer = "locs.gb", driver = "KML")
```

### Reading KML

```{r}
newmap <- readOGR("locsgb.kml", layer = "locs.gb")
```

### Save as shapefile

```{r}
writePointsShape(locs.gb, "locsgb")
```

### Reading shapefiles

```{r}
gb.shape <- readShapePoints("locsgb.shp")
plot(gb.shape)
```

Use `readShapePoly` to read polygon shapefiles, and `readShapeLines` to read polylines. See also `shapefile` in `raster` package.   

## 3.9 Changing projection of spatial vector data

`spTransform` (package `sp`) will do the projection as long as the original and new projection are correctly specified.

### Projecting point dataset

To illustrate, let's project the dataframe with Laurus nobilis coordinates that we obtained above:

```{r}
summary(locs)
```

The original coordinates are in lat lon format. Let's define the new desired projection:

Lambert Azimuthal Equal Area in this case 

(look up parameters at [http://spatialreference.org](http://spatialreference.org))

```{r}
crs.laea <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") # Lambert Azimuthal Equal Area
locs.laea <- spTransform(locs, crs.laea)    # spTransform makes the projection
```

### Projecting shapefile of countries

```{r}
plot(countriesLow)    # countries map in geographical projection
country.laea <- spTransform(countriesLow, crs.laea)  # project
```

Let's plot this:

```{r}
plot(locs.laea, pch = 20, col = "steelblue")
plot(country.laea, add = TRUE)
# define spatial limits for plotting
plot(locs.laea, pch = 20, col = "steelblue", xlim = c(1800000, 3900000), 
     ylim = c(1000000, 3000000))
plot(country.laea, add = TRUE)
```

# 4. USING RASTER (GRID) DATA

## 4.1 Downloading raster climate data from internet

The `getData` function from the `dismo` package will easily retrieve climate data, elevation, administrative boundaries, etc. Check also the excellent [rWBclimate package](http://ropensci.org/packages/rwbclimate.html) by rOpenSci with additional functionality. 

```{r}
tmin <- getData("worldclim", var = "tmin", res = 10)   
# this will download global data on minimum temperature at 10' resolution
```

## 4.2 Loading a raster layer

```{r}
tmin1 <- raster(paste(getwd(), "/wc10/tmin1.bil", sep = ""))   # Tmin for January
```

Easy! The `raster` function reads many different formats, including Arc ASCII grids or netcdf files (see raster help). And values are stored on disk instead of memory! (useful for large rasters) 

```{r}
fromDisk(tmin1)
```

Let's examine the raster layer:

```{r}
tmin1 <- tmin1/10    # Worldclim temperature data come in decimal degrees 
tmin1    # look at the info
plot(tmin1)
```

## 4.3 Creating a raster stack

A raster stack is collection of many raster layers with the same projection, spatial extent and resolution. Let's collect several raster files from disk and read them as a single raster stack:

```{r message=FALSE, warning=FALSE}
file.remove(paste(getwd(), "/wc10/", "tmin_10m_bil.zip", sep = ""))
list.ras <- mixedsort(list.files(paste(getwd(), "/wc10/", sep = ""), full.names = TRUE,
                                 pattern = ".bil"))
list.ras # I have just collected a list of the files containing monthly temperature values
tmin.all <- stack(list.ras)
tmin.all
tmin.all <- tmin.all/10
plot(tmin.all)
```

## 4.4 Raster bricks

A rasterbrick is similar to a raster stack (i.e. multiple layers with the same extent and resolution), but all the data must be stored in a single file on disk.

```{r}
tmin.brick <- brick(tmin.all)   # creates rasterbrick
```

## 4.5 Crop rasters

Crop raster manually (drawing region of interest):

```{r eval=FALSE}
plot(tmin1)
newext <- drawExtent()    # click twice on the map to select the region of interest
tmin1.c <- crop(tmin1, newext)
plot(tmin1.c)
```

Alternatively, provide coordinates for the limits of the region of interest:

```{r}
newext <- c(-10, 10, 30, 50)   
tmin1.c <- crop(tmin1, newext)
plot(tmin1.c)
tmin.all.c <- crop(tmin.all, newext)
plot(tmin.all.c)
```

## 4.6 Define spatial projection of the rasters

```{r}
crs.geo    # defined above
projection(tmin1.c) <- crs.geo
projection(tmin.all.c) <- crs.geo
tmin1.c    # notice info at coord.ref.
```

## 4.7 Changing projection

Use `projectRaster` function:

```{r}
tmin1.proj <- projectRaster(tmin1.c, crs="+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +a=6378137 +b=6378137 +units=m +no_defs")   # can also use a template raster, see ?projectRaster
tmin1.proj   # notice info at coord.ref.
plot(tmin1.proj)
```

## 4.8 Plotting raster data

Different plotting functions:

```{r}
histogram(tmin1.c)
pairs(tmin.all.c)
persp(tmin1.c)
contour(tmin1.c)
contourplot(tmin1.c)
levelplot(tmin1.c)
plot3D(tmin1.c)
bwplot(tmin.all.c)
densityplot(tmin1.c)
```

## 4.9 Spatial autocorrelation

```{r}
Moran(tmin1.c)    # global Moran's I
tmin1.Moran <- MoranLocal(tmin1.c)
plot(tmin1.Moran)
```

## 4.10 Extract values from raster

Use `extract` function:

```{r}
head(locs)    # we'll obtain tmin values for our points
projection(tmin1) <- crs.geo
locs$tmin1 <- extract(tmin1, locs)    
# raster values are incorporated to the dataframe
head(locs)
```

You can also extract values for a given region instead of the whole raster:

```{r eval=FALSE}
plot(tmin1.c)
reg.clim <- extract(tmin1.c, drawExtent())  
# click twice to draw extent of the region of interest 
summary(reg.clim)
```

Using `rasterToPoints`:

```{r}
# rasterToPoints
tminvals <- rasterToPoints(tmin1.c)
head(tminvals)
```

And also, the `click` function will get values from particular locations in the map

```{r eval=FALSE}
plot(tmin1.c)
click(tmin1.c, n=3)   # click n times in the map to get values
```

## 4.11 Rasterize points, lines or polygons

```{r}
locs2ras <- rasterize(locs.gb, tmin1, field = rep(1, nrow(locs.gb)))
locs2ras
plot(locs2ras, xlim = c(-10, 10), ylim = c(45, 60), legend = FALSE)
data(wrld_simpl)
plot(wrld_simpl, add = TRUE)
```

## 4.12 Changing raster resolution

Use `aggregate` function:

```{r}
tmin1.lowres <- aggregate(tmin1.c, fact=2, fun=mean)
tmin1.lowres
tmin1.c     # compare
par(mfcol = c(1,2))
plot(tmin1.c, main = "original")
plot(tmin1.lowres, main = "low resolution")
par(mfcol = c(1,1))
```

## 4.13 Spline interpolation

```{r message=FALSE, warning=FALSE}
xy <- data.frame(xyFromCell(tmin1.lowres, 1:ncell(tmin1.lowres))) # get raster cell coordinates
head(xy)
vals <- getValues(tmin1.lowres)
spline <- Tps(xy, vals)    # thin plate spline
intras <- interpolate(tmin1.c, spline)
intras    # note new resolution
plot(intras)  
intras <- mask(intras, tmin1.c)   # mask to land areas only
plot(intras)
title("Interpolated raster")
```

## 4.14 Setting all rasters to the same extent, projection and resolution all in one

See `spatial_sync_raster` function from `spatial.tools` package.

## 4.15 Elevations, slope, aspect, etc

Download elevation data from internet:

```{r}
elevation <- getData('alt', country = 'ESP')
```

Some quick maps:

```{r}
x <- terrain(elevation, opt = c('slope', 'aspect'), unit = 'degrees')
plot(x)
slope <- terrain(elevation, opt = 'slope')
aspect <- terrain(elevation, opt = 'aspect')
hill <- hillShade(slope, aspect, 40, 270)
plot(hill, col = grey(0:100/100), legend = FALSE, main = 'Spain')
plot(elevation, col = rainbow(25, alpha = 0.35), add = TRUE)
```

## 4.16 Saving and exporting raster data

Saving raster to file:

```{r}
writeRaster(tmin1.c, filename="tmin1.c.grd")   
writeRaster(tmin.all.c, filename="tmin.all.grd")
```

`writeRaster` can export to many different file types, see help.

Exporting to KML (Google Earth)

```{r}
tmin1.c <- raster(tmin.all.c, 1)
KML(tmin1.c, file="tmin1.kml")  
KML(tmin.all.c)     # can export multiple layers

```

# 5. SPATIAL STATISTICS (just a glance)

## 5.1 Point pattern analysis

Some useful packages:

```{r message=FALSE}
library(spatial)
library(spatstat)
library(spatgraphs)
library(ecespa)    # ecological focus
```

See [CRAN Spatial Task View](http://cran.r-project.org/web/views/Spatial.html).

Let's do a quick example with Ripley's K function:

```{r}
data(fig1)
plot(fig1)    # point pattern
data(Helianthemum)
cosa12 <- K1K2(Helianthemum, j="deadpl", i="survpl", r=seq(0,200,le=201),
               nsim=99, nrank=1, correction="isotropic")
plot(cosa12$k1k2, lty=c(2, 1, 2), col=c(2, 1, 2), xlim=c(0, 200),
     main= "survival- death",ylab=expression(K[1]-K[2]), legend=FALSE)
```

## 5.2 Geostatistics

Some useful packages:

```{r message=FALSE, eval=FALSE}
library(gstat)
library(geoR)
library(akima)   # for spline interpolation
library(spdep)   # dealing with spatial dependence
```

See [CRAN Spatial Task View](http://cran.r-project.org/web/views/Spatial.html).

# 6. INTERACTING WITH OTHER GIS

```{r message=F, eval=F}
library(spgrass6)   # GRASS
library(RPyGeo)     # ArcGis (Python)
library(RSAGA)      # SAGA
library(spsextante) # Sextante 
```

# 7. OTHER USEFUL PACKAGES

```{r message=FALSE, eval=FALSE}
library(Metadata)    # automatically collates data from online GIS datasets (land cover, pop density, etc) for a given set of coordinates
#library(GeoXp)    # Interactive exploratory spatial data analysis

example(columbus)
histomap(columbus, "CRIME")

library(maptools)
# readGPS 

library(rangeMapper)    # plotting species distributions, richness and traits

# Species Distribution Modelling
library(dismo)
library(biomod2)
library(SDMTools)

library(BioCalc)   # computes 19 bioclimatic variables from monthly climatic values (tmin, tmax, prec)
```

# 8. TO LEARN MORE

- [ASDAR book](http://www.asdar-book.org/)

- Packages help and vignettes, especially

http://cran.r-project.org/web/packages/raster/vignettes/Raster.pdf   

http://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf   

http://cran.r-project.org/web/packages/sp/vignettes/sp.pdf   

- [CRAN Task View: Analysis of Spatial Data](http://cran.r-project.org/web/views/Spatial.html)

- [Introduction to Spatial Data and ggplot2](http://rpubs.com/RobinLovelace/intro-spatial)

- [R spatial tips](http://spatial.ly/category/r-spatial-data-hints/)

- [R wiki: tips for spatial data](http://rwiki.sciviews.org/doku.php?id=tips:spatial-data&s=spatial)

- [Spatial analysis in R](http://www.maths.lancs.ac.uk/~rowlings/Teaching/Sheffield2013/index.html)

- [Displaying time series, spatial and space-time data with R](http://oscarperpinan.github.io/spacetime-vis/)

- [Notes on Spatial Data Operations in R](https://dl.dropboxusercontent.com/u/9577903/broomspatial.pdf)

- [Analysing spatial point patterns in R](http://www.csiro.au/resources/pf16h)

- [Spatial data in R](http://science.nature.nps.gov/im/datamgmt/statistics/r/advanced/Spatial.cfm)

- [NCEAS Geospatial use cases](http://www.nceas.ucsb.edu/scicomp/usecases)

- [Spatial Analyst](http://spatial-analyst.net)

- [Making maps with R](http://www.molecularecologist.com/2012/09/making-maps-with-r/)

- [The Visual Raster Cheat Sheet](http://www.rpubs.com/etiennebr/visualraster)

- [R-SIG-Geo mailing list](https://stat.ethz.ch/mailman/listinfo/R-SIG-Geo)

- [Geospatial data processing and analysis in R (slides)](http://rpubs.com/ajlyons/rspatialdata)

- [rMaps](http://rmaps.github.io/)
